---
title: "Number_Trees_BI_nDOM"
format: html
knitr:
  opts_chunk: 
    eval: false
editor: visual
---

## Compare number of trees in BI plots: detected with remote sensing vs measured in field

```{r}
library(here)
library(sf)
library(dplyr)
library(terra)
library(ggplot2)
library(foreign)
```

Load BI plots, BI trees and trees detected in ndom (created in previous script).

```{r}
plots_bi <- st_read(here("R_session", "data", "buffer13_bi_mittelpunkt_solling_2023.gpkg"))

trees_bi <- st_read(here("R_session", "data", "bi_baeume_vorrat_solling_2023.gpkg"))

trees_ndom <- st_read(here("R_session", "data", "ttops_bi_ndom.gpkg"))
```

Load table with median slope in the BI plot and bind with plots BI table

```{r}
slope_bi_plot <- read.dbf(here("R_session", "data", "bi_plot_median_slope.dbf"))
plots_bi <- bind_cols(plots_bi, slope_bi_plot %>% select(X_median))
rm(slope_bi_plot)
```

Count how many bi trees inside plot

```{r}
intersection_bi_trees_plots <- st_intersection(x = plots_bi, y = trees_bi)

count_bi_trees_plot <- intersection_bi_trees_plots %>% 
    add_count(KSPNR, name = "n_trees_bi") %>% 
    group_by(KSPNR) %>% 
    slice(n()) %>% 
    ungroup() 
```

Optional: Trees_bi 2638 observations, intersection_bi_trees_plots 2606 observations, check reason from this difference

```{r}
#st_erase = function(x, y) st_difference(x, st_union(st_combine(y)))

#difference <- st_erase(trees_bi , intersection_bi_trees_plots)

#writeVector(vect(difference), here("output", "trees_not_bi.gpkg"), overwrite=TRUE)
```

Count how many ndom trees inside plot

```{r}
intersection_ndom_trees_plots <- st_intersection(x = plots_bi, y = trees_ndom)

count_ndom_trees_plot <- intersection_ndom_trees_plots %>% 
    add_count(KSPNR, name = "n_trees_ndom") %>% 
    group_by(KSPNR) %>% 
    slice(n()) %>% 
    ungroup() 
```

Merge table with number trees from ndom and table with number trees from BI

```{r}
count_trees_bi_ndom <- merge(as.data.frame(count_ndom_trees_plot), as.data.frame(count_bi_trees_plot), by = "KSPNR")
```

Plot to compare the number of trees in each plot taking slope of plot into account

```{r}
x <- ggplot(count_trees_bi_ndom, aes(x=n_trees_bi, y=n_trees_ndom,        colour=X_median.x)) + 
              xlim(0,30) + ylim(0,30) +
              geom_abline(slope=1) +
              geom_point() +
              scale_colour_gradientn(colours=rainbow(4)) +
              labs(title = "Number of trees in each plot measured in BI vs detected with nDOM", 
              x = "Number of trees measured in BI",
              y = "Number of trees detected in nDOM",
              color = "Terrain slope [degree]") +
              theme(plot.title = element_text(hjust = 0.5))
```

What could be the reasons for the difference?
